Insights of Even-point Multi-loop Unitarity: Exponentiation, Anomalies and Evanescence

### 2023-06-08 eventPointTwoLoop.tex
Intro and sec 2-3 updated
Added more to the intro, alog with ampliutde review, double copy + CKD review, spinnor helicity and D-dim review, even point theory review (paartial), moved somoething things aroud to fix logiccal structure

\documentclass[12pt,letter]{article}
\usepackage{jheppub}
\usepackage{amsmath,amssymb}
\usepackage[dvipsnames]{xcolor}
%\usepackage{hyperref}
\usepackage{xspace}
\usepackage{ifdraft}
\usepackage{epstopdf}
\usepackage{slashed}
\usepackage{diagbox}
\usepackage{colortbl}
\usepackage{bbold}
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{stackengine}
\def\RA{\rlap{\scalebox{1.6}{$\leftrightarrow$}}}
\def\DA{\smash{\bclap{\scalebox{1.6}{$\downarrow$}}}}
\def\mystrut{\rule[-2ex]{0ex}{6ex}}

@@

\newcommand{\HexTri}{ {
\begin{tikzpicture}[baseline=2]
\begin{feynman}
\vertex (a1) at (1.73*.6, 1*.6) {};
\vertex (a2) at (0*.6, 2*.6) {};
\vertex (a3) at (-1.73*.6, 1*.6) {};
\vertex [dot,scale=3.6*.6](mid1) at (.865*.6, .5*.6){};
\vertex [dot,scale=3*.6,hred] (mid1a) at (.865*.6, .5*.6){};
\vertex [dot,scale=3.6*.6](mid2) at (0*.6, 1*.6){};
\vertex [dot,scale=3*.6,hred] (mid2a) at (0*.6, 1*.6){};
\vertex [dot,scale=3.6*.6](mid3) at (-.865*.6, .5*.6){};
\vertex [dot,scale=3*.6,hred] (mid3a) at (-.865*.6, .5*.6){};
\diagram{
(mid1) --[ ultra thick](a1),
(mid2) --[ ultra thick](a2),
(mid3) --[ ultra thick](a3),
(mid1) --[ ultra thick](mid2),
(mid3) --[ ultra thick](mid2),
(mid3) --[ ultra thick](mid1),
};
\end{feynman}
\end{tikzpicture}
}
}

\newcommand{\maxHex}{ {
\begin{tikzpicture}[baseline=2]
\begin{feynman}
\vertex (a1) at (1.73*.6, 1*.6) {};
\vertex (a2) at (0*.6, 2*.6) {};
\vertex (a3) at (-1.73*.6, 1*.6) {};
\vertex (a4) at (-1.73*.6, -1*.6) {};
\vertex (a5) at (0*.6, -2*.6) {};
\vertex (a6) at (1.73*.6, -1*.6) {};
\vertex [dot,scale=3.6*.6](mid1) at (.865*.6, .5*.6){};
\vertex [dot,scale=3*.6,hblue] (mid1a) at (.865*.6, .5*.6){};
\vertex [dot,scale=3.6*.6](mid2) at (0*.6, 1*.6){};
\vertex [dot,scale=3*.6,hblue] (mid2a) at (0*.6, 1*.6){};
\vertex [dot,scale=3.6*.6](mid3) at (-.865*.6, .5*.6){};
\vertex [dot,scale=3*.6,hblue] (mid3a) at (-.865*.6, .5*.6){};
\vertex [dot,scale=3.6*.6](mid4) at (-.865*.6, -.5*.6){};
\vertex [dot,scale=3*.6,hblue] (mid4a) at (-.865*.6, -.5*.6){};
\vertex [dot,scale=3.6*.6](mid5) at (0*.6, -1*.6){};
\vertex [dot,scale=3*.6,hblue] (mid5a) at (0*.6, -1*.6){};
\vertex [dot,scale=3.6*.6](mid6) at (.865*.6, -.5*.6){};
\vertex [dot,scale=3*.6,hblue] (mid6a) at (.865*.6, -.5*.6){};
\diagram{
(mid1) --[ ultra thick](a1),
(mid2) --[ ultra thick](a2),
(mid3) --[ ultra thick](a3),
(mid4) --[ ultra thick](a4),
(mid5) --[ ultra thick](a5),
(mid6) --[ ultra thick](a6),
(mid1) --[ ultra thick](mid2),
(mid3) --[ ultra thick](mid2),
(mid3) --[ ultra thick](mid4),
(mid5) --[ ultra thick](mid4),
(mid5) --[ ultra thick](mid6),
(mid1) --[ ultra thick](mid6),
};
\end{feynman}
\end{tikzpicture}
}
}

@@

\newcommand{\Hspin}{ {
\begin{tikzpicture}[baseline=-3]
\begin{feynman}
\vertex (a1) at (-.7,-.7) {};
\vertex (a2) at (-.7,.7) {};
\vertex (a3) at (.7,.7) {};
\vertex (a4) at (.7,-.7) {};
\vertex (mid1) at (-.35,0);
\vertex (mid2) at (.35,0);
\vertex (mid1a) at (-.35,.15);
\vertex (mid2a) at (.35,.15);
\vertex (mid1b) at (-.35,-.15);
\vertex (mid2b) at (.35,-.15);
\diagram{
(mid1) -- [photon,ultra thick] (a1);
(mid1) -- [photon,ultra thick] (a2);
(mid2) -- [photon,ultra thick] (a3);
(mid2) -- [photon,ultra thick] (a4);
(mid1) -- [photon,ultra thick,hred] (mid2);
(mid1a) -- [ photon,ultra thick,hred](mid2a);
(mid1b) -- [ photon,ultra thick,hred] (mid2b);

@@

@@

\title{Even-point Multi-loop Unitarity and its Applications: Exponentiation, Anomalies and Evanescence}

\abstract{
We identify novel  structure in newly computed multi-loop amplitudes and quantum actions for a special class of effective field theories which includes both the nonlinear sigma model (NLSM) and its double-copy gauge theories such as Born-Infeld and supersymmetric generalizations.  We exploit special properties of these Even-point theories to generalize the direct extraction of Feynman integral coefficients using $D$-dimensional unitarity cuts to the multi-loop case for the first time.  Doing so, we find that the leading IR divergence of NLSM amplitudes exponentiates. Second, we find that the counterterms required to restore U(1) behavior at loop-level in the Born-Infeld theory can be constructed via a symmetric-structure double-copy.  We demonstrate that the one-minus $(-+++)$ two-loop amplitude vanishes upon the introduction of an evanescent operator in the $D$-dimensional description of Born-Infeld theory. In addition to these pure photon counterterms, we verify through explicit calculation that the anomalous matrix elements that violates $U(1)$ duality invariance can be alternatvely cancelled by summing over internal $\mathcal{N}=4$ DBIVA superfields. Finally we find that $\mathcal{N}=4$ Dirac-Born-Infeld-Volkov-Akulov (DBIVA) amplitudes permit double-copy construction through two-loop order. This is shown by reproducing our unitarity based result with a loop-level double copy between $\mathcal{N}=4$ super-Yang-Mills numerators and our two-loop NLSM amplitudes. This result provides strong evidence for the existence of a color-dual representations for NLSM beyond one-loop. We conclude with an overview of how $D$-dimensional four-photon counterterms can be constructed in generality with the symmetric-structure double-copy, and outline a convenient way of counting evanescent operators at higher orders using Hilbert series as generating functions. }

@@

\section{Introduction}\label{sec:intro}
\dj{Probably a more basic introduction for people who don't know what generalized unitarity is, or double copy, poor souls.}
\dj{Emphasize one advantage of unitarity methods at 1-loop is the ability to integrate by doing cuts -- i.e. know exactly the coefficients of each basis integral.  We show for a very special class of theories we can morally do exactly the same at higher loops -- why? even points,recursion blah blah balh.  In sect 2 we review all the old.  In sect 3 we lay out Even-point Multi-loop Unitarity in detail what properties we rely on and how we exploit them blah blah.  }

Here we study even-point (EP) effective field theory (EFT) at the multi-loop level using a combination of generalized unitarity \cite{UnitarityMethod, Fusing, BDKUniarityReview} and the double-copy construction \cite{BCJ, BCJLoop}. The EFTs that we consider in this work are the nonlinear sigma model (NLSM) and a related family of gauge theories involving Born-Infeld theory and supersymmetric generalizations known as Dirac-Born-Infeld-Volkov-Akulov (DBIVA) theories. These families are known at tree-level to be double-copies between NLSM and Yang-Mills theories with varying amounts of supersymmetry. However, the methods we use can be similarly used for perturbative calculations in other EP effective field theories.

The motivation for this work is three-fold. First, perturbative calculations in NLSM and DBIVA effective field theories are significantly simpler than the typical multi-loop calculations of renormalizable theories, like quantum-chromodynamics (QCD) and super-Yang-Mills (sYM). As we will show, the multi-loop amplitudes for these theories allow us to recycle many of the $D$-dimensional integration tools that are so powerful at one-loop order. This allows us to compute a large catalog of $D$-dimensional two-loop amplitudes, which provides fertile ground for cultivating insights about the multi-loop structure of effective field theories. The perturbative depth of our calculations sheds light on the exponential structure of IR divergences for NLSM amplitudes through two-loops, the anomalous behavior $U(1)$ duality invariance through two-loop order in DBIVA theories, as well as the emergence of evanescent operators relevant for anomaly cancellation in pure Born-Infeld theory. We believe these surprisingly rich physical structures could serve as a theoretical laboratory for future studies of the interplay between effective field theory operators and perturbative calculations in quantum field theory.

The second motivation is that very little is known about color-kinematics duality at multi-loop level for generic models in the web of theories \cite{BCJreview}. For a comprehensive review of color-dual representations, see refs. \cite{BCJreview, Bern:2022wqg, Adamo:2022dcm} and references therein. While there has been tremendous success of applying color-kinematics duality to $\mathcal{N}=4$ super-Yang-Mills, for which color-dual integrands are known through four-point four-loop \cite{Bern:2012uf, Neq44np, GravityFour}, five-point through three-loops, and to sEven-point at one-loop \cite{Bjerrum-Bohr:2013iza,Edison:2020uzf,Edison:2022jln}, there are many obstructions for generic gauge/gravity theories. Presently, the state-of-the-art for nonlinear sigma model (NLSM) color-dual numerators come from the $XY\!Z$ model of Cheung and Shen \cite{Cheung2016prv}, while $D$-dimensional integrands for pure Yang-Mills have only been identified through five-point one-loop \cite{He:2017spx}. At present, there are no known $D$-dimensional representations for either of these theories at two-loop that globally manifest the duality between color and kinematics, despite attempts in the literature \cite{OneTwoLoopPureYMBCJ, Mogull:2015adi, Bern:2015ooa, Geyer:2019hnn}. Furthermore, similar\dj{?? I though they had no problems with 2-loops.} bottlenecks exists for less-than-maximal $\mathcal{N}\leq 2$ sYM \cite{Johansson:2017bfl}. While there have been a number of recent developments in constructing manifestly color-dual Feynman rules \cite{Chen:2019ywi,Chen:2021chy,Brandhuber:2021bsf,Cheung:2021zvb,Ben-Shahar:2021zww,Cheung:2022mix, Ben-Shahar:2022ixa}, a precise definition the kinematic algebra off-shell remains elusive. Considering the known tree-level double-copy relationships between NLSM and DBIVA and higher-derivative corrections \cite{Cachazo:2014xea,Carrasco:2016ldy}, all of the amplitudes we compute should in principle participate in a double-copy formulation of our results. At the very least, the amplitudes we compute will serve as important checks on future breakthroughs in multi-loop studies of color-dual representations. In addition to traditional double-copy construction, we find that the quantum effective actions generated by these loop effect of this paper permit a rather compact construction in terms of \textit{symmetric-structure double-copy}, introduced in a recent work by the authors~\cite{Carrasco:2022jxn}.

@@

\dj{To approach these problems we generalize the by now standard one-loop unitarity approach of Forde to higher loops exploiting a special feature of our theories blah blah.}

The outline of the paper is as follows: in \sect{sec:Review}, we will provide a review of the on-shell methods and integration techniques  needed to probe the multi-loop physics studied in this paper. Then in \sect{sec:EMU}, we introduce the method of Even-point Multiloop Unitarity (EMU) and compute the two-loop tensor integrals needed in this work. In \sect{sec:Loops}, we present our results, beginning with a warm-up calculation in \sect{sec:NLSMU} where we construct the integrands for NLSM through NNLO in the effective coupling, and compute the fully integrated amplitudes. We show that in $D=2-2\epsilon$, the leading IR divergences of the theory exponentiate. With the scaffolding of $D$-dimensional integration in hand, we then compute two-loop amplitudes for $\mathcal{N}=4$ DBIVA theory and pure-photon Born-Infeld theory in \sect{sec:DBIU}. There we compute the anomalies present beyond one-loop order.  Then, using the NLSM integrands of \sect{sec:NLSMU}, we perform a multi-loop double-copy to $\mathcal{N}=4$ DBIVA observables with the color-dual basis numerators available in the literature in \sect{sec:DBIvDC}. After computing the catalog of two-loop amplitudes, we study the construction of quantum effective actions in \sect{sec:Actions} as they relate to the anomalous matrix elements present in two-loop Born-Infeld amplitudes. We demonstrate in \sect{sec:Anomalies} that the one-minus anomaly at two-loop requires the introduction of an evanescent operator at $\mathcal{O}(\alpha'^4)$. Then in \sect{sec:ActionDC} we take the opportunity to discuss the application of symmetric-structure double copy to cancel these anomalies, along with their relationship to higher-spin modes as studied in a recent work by the authors. Finally, in \sect{sec:EOpsCounting} we lay out a Hilbert series framework for counting evanescent operators at general orders in mass-dimension. To conclude, in \sect{sec:Discussion} we discuss many directions of future work and summarize the insights gained from this study.

\section{Review} 
\label{sec:Review}
\subsection{Color-dressed and ordered amplitudes}
\label{subsec:AmpReview}
\dj{2.1 amplitudes and nomenclature (integrand, numerators, denominators, symmetry factors, color-ordered amplitudes)., e.g. when we use a curly A or M what does it mean and why. Also throughout the paper we should standardize, $\{\cal{O},{\cal{M}}\}\to\cal{A}$ .  I guess we use straight $\epsilon$ for dimensions, and $\varepsilon$ for polarizations.  We can explain polarizations in a 4D vs D-dim part. I would do standard choices so normal people (like us) can skip this section but if they hand it to some rando QCD person or string person to referee they can figure out what we're talking about. I don't know, between a page or two here maybe.}

\subsection{Color-Kinematics Duality and Double-Copy Construction}
\label{subsec:DCReview}

\dj{color-kinematics and double-copy (this could literally be the standard paragraph or so, trivalent graphs, color-weights, maybe a little longer than usual because we're also introducing $d's$)}

\subsection{4D Spinor Helicity vs $D$-dimensions}
\label{subsec:4DandDDReview}
\dj{4D (spinor helicity) vs arb D dimensions and  formal polarizations. (maybe half a page).  }

\subsection{Even-point Effective Field Theories}
\label{subsec:EPEFTReview}
\dj{EP theories and relevant effective actions, hilbert series, state of the art of what has been calculated in each theory. (move some of the introduction discussion here, but this could be a couple of pages.)}

We here provide a brief review of the types of models studied in this work, namely  NLSM and DBIVA.

\dj{copied from later}
The Lagrangian for the \textit{nonlinear sigma model} (NLSM) can be expressed in terms of the chiral current, $j_\mu = U^\dagger \partial_\mu U$, as follows,
\be
\mathcal{L}^{\text{NLSM}}= \frac{1}{2}\text{tr}[(\partial_\mu U)^\dagger (\partial ^\mu U)] = \frac{1}{2}\text{tr}\left[\frac{(\partial_\mu\pi )( \partial^\mu \pi) }{(1-f_\pi^{-2}\pi^2)^2}\right]
\ee
where the trace is over the color indices of the gauge group, $\pi \equiv \pi^a T^a$, and on the right hand side we have applied the Cayley parameterization to the $SU(N_c)$ group elements, $U= \frac{1+\pi/f_\pi}{1-\pi/f_\pi}$, where $f_\pi$ is the dimensionful pion decay width. As noted in the preamble to this section, to construct the integrand, we will impose a spanning set of generalized unitarity cuts that appear in a minimal color basis. This procedure will provide the scaffolding for the colorless amplitudes of interest in the latter part of the paper.  This lagrangian is the same independent of what dimension.

In contrast, maximally supersymmetric Born-Infeld comes from adding the maximal amount of supersymmetric fermionic and scalar matter to Born-Infeld.  The Born-Infeld Lagrangian, which describes the pure vector sector of DBIVA theories, can be expressed concisely below:
\begin{equation}
 \label{biLag}
  \alpha'^2\mathcal{L}^{\text{BI}} = 1-\sqrt{1+\frac{(\alpha')^2}{2} (F_{\mu\nu}F^{\mu\nu})-\frac{(\alpha')^4}{16}(F_{\mu\nu}\tilde{F}^{\mu\nu})^2}\,.
\end{equation}
Throughout the remainder of the paper, we will set $\alpha'=1$ in the amplitude expressions.

\dj{Prolly add at least fermionic actions so maybe N=1 in 10D, and mention dimensional reduction to get maximally susy in lower D?}
\begin{equation}
 \label{vaLag}
  \alpha'^2\mathcal{L}^{\text{VA}} =\ldots
  \,.
\end{equation}

>>>>

\documentclass[12pt,letter]{article}
\usepackage{jheppub}
\usepackage{amsmath,amssymb}
\usepackage[dvipsnames]{xcolor}
%\usepackage{hyperref}
\usepackage{xspace}
\usepackage{ifdraft}
\usepackage{epstopdf}
\usepackage{slashed}
\usepackage{diagbox}
\usepackage{colortbl}
\usepackage{shuffle}
\usepackage{bbold}
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{stackengine}
\def\RA{\rlap{\scalebox{1.6}{$\leftrightarrow$}}}
\def\DA{\smash{\bclap{\scalebox{1.6}{$\downarrow$}}}}
\def\mystrut{\rule[-2ex]{0ex}{6ex}}

@@

\newcommand{\HexTri}{ {
\begin{tikzpicture}[baseline=2]
\begin{feynman}
\vertex (a1) at (1.73*.6, 1*.6) {};
\vertex (a2) at (0*.6, 2*.6) {};
\vertex (a3) at (-1.73*.6, 1*.6) {};
\vertex [dot,scale=3.6*.6](mid1) at (.865*.6, .5*.6){};
\vertex [dot,scale=3*.6,hred] (mid1A) at (.865*.6, .5*.6){};
\vertex [dot,scale=3.6*.6](mid2) at (0*.6, 1*.6){};
\vertex [dot,scale=3*.6,hred] (mid2a) at (0*.6, 1*.6){};
\vertex [dot,scale=3.6*.6](mid3) at (-.865*.6, .5*.6){};
\vertex [dot,scale=3*.6,hred] (mid3a) at (-.865*.6, .5*.6){};
\diagram{
(mid1) --[ ultra thick](a1),
(mid2) --[ ultra thick](a2),
(mid3) --[ ultra thick](a3),
(mid1) --[ ultra thick](mid2),
(mid3) --[ ultra thick](mid2),
(mid3) --[ ultra thick](mid1),
};
\end{feynman}
\end{tikzpicture}
}
}

\newcommand{\maxHex}{ {
\begin{tikzpicture}[baseline=2]
\begin{feynman}
\vertex (a1) at (1.73*.6, 1*.6) {};
\vertex (a2) at (0*.6, 2*.6) {};
\vertex (a3) at (-1.73*.6, 1*.6) {};
\vertex (a4) at (-1.73*.6, -1*.6) {};
\vertex (a5) at (0*.6, -2*.6) {};
\vertex (a6) at (1.73*.6, -1*.6) {};
\vertex [dot,scale=3.6*.6](mid1) at (.865*.6, .5*.6){};
\vertex [dot,scale=3*.6,hblue] (mid1A) at (.865*.6, .5*.6){};
\vertex [dot,scale=3.6*.6](mid2) at (0*.6, 1*.6){};
\vertex [dot,scale=3*.6,hblue] (mid2a) at (0*.6, 1*.6){};
\vertex [dot,scale=3.6*.6](mid3) at (-.865*.6, .5*.6){};
\vertex [dot,scale=3*.6,hblue] (mid3a) at (-.865*.6, .5*.6){};
\vertex [dot,scale=3.6*.6](mid4) at (-.865*.6, -.5*.6){};
\vertex [dot,scale=3*.6,hblue] (mid4a) at (-.865*.6, -.5*.6){};
\vertex [dot,scale=3.6*.6](mid5) at (0*.6, -1*.6){};
\vertex [dot,scale=3*.6,hblue] (mid5a) at (0*.6, -1*.6){};
\vertex [dot,scale=3.6*.6](mid6) at (.865*.6, -.5*.6){};
\vertex [dot,scale=3*.6,hblue] (mid6a) at (.865*.6, -.5*.6){};
\diagram{
(mid1) --[ ultra thick](a1),
(mid2) --[ ultra thick](a2),
(mid3) --[ ultra thick](a3),
(mid4) --[ ultra thick](a4),
(mid5) --[ ultra thick](a5),
(mid6) --[ ultra thick](a6),
(mid1) --[ ultra thick](mid2),
(mid3) --[ ultra thick](mid2),
(mid3) --[ ultra thick](mid4),
(mid5) --[ ultra thick](mid4),
(mid5) --[ ultra thick](mid6),
(mid1) --[ ultra thick](mid6),
};
\end{feynman}
\end{tikzpicture}
}
}

@@

\newcommand{\Hspin}{ {
\begin{tikzpicture}[baseline=-3]
\begin{feynman}
\vertex (a1) at (-.7,-.7) {};
\vertex (a2) at (-.7,.7) {};
\vertex (a3) at (.7,.7) {};
\vertex (a4) at (.7,-.7) {};
\vertex (mid1) at (-.35,0);
\vertex (mid2) at (.35,0);
\vertex (mid1A) at (-.35,.15);
\vertex (mid2a) at (.35,.15);
\vertex (mid1B) at (-.35,-.15);
\vertex (mid2b) at (.35,-.15);
\diagram{
(mid1) -- [photon,ultra thick] (a1);
(mid1) -- [photon,ultra thick] (a2);
(mid2) -- [photon,ultra thick] (a3);
(mid2) -- [photon,ultra thick] (a4);
(mid1) -- [photon,ultra thick,hred] (mid2);
(mid1A) -- [ photon,ultra thick,hred](mid2a);
(mid1B) -- [ photon,ultra thick,hred] (mid2b);

@@

\newcommand{\threeTree}[3]{ {
\begin{tikzpicture}[baseline=7]
\begin{feynman}
\vertex (a1) at (-1.4,0){$i$} ;
\vertex (a2) at (-.4,1){#1} ;
\vertex (a3) at (.4,1){#2} ;
\vertex (a4) at (1.4,0){#3};
\vertex (mid1) at (-.4,0);
\vertex (mid2) at (.4, -0) ;
\diagram{
(mid1) --[ultra thick,](a1),
(mid1) --[ultra thick,](a2),
(mid1) --[ultra thick,](mid2),
(mid2) --[ultra thick,](a3),
(mid2) --[ultra thick,](a4),
};
\end{feynman}
\end{tikzpicture}
}
}

@@

\title{Insights of Even-point Multi-loop Unitarity: Exponentiation, Anomalies and Evanescence}

\abstract{
We identify novel  structure in newly computed multi-loop amplitudes and quantum actions for even-point effective field theories, including both the nonlinear sigma model (NLSM) and double-copy gauge theories such as Born-Infeld and its supersymmetric generalizations.  We exploit special properties of these even-point theories to generalize the direct extraction of Feynman integral coefficients using $D$-dimensional unitarity cuts to the multi-loop case for the first time. In doing so, we find that the leading IR divergence of NLSM amplitudes exponentiates. We then systematically compute the two-loop anomalous behavior of Born-Infeld, and find that the counterterms required to restore $U(1)$ behavior at loop-level in the Born-Infeld theory can be constructed via a symmetric-structure double-copy.  We also demonstrate that the one-minus $(-+++)$ two-loop amplitude vanishes upon the introduction of an evanescent operator in the $D$-dimensional description of Born-Infeld theory. In addition to these pure photon counterterms, we verify through explicit calculation that the anomalous matrix elements that violates $U(1)$ duality invariance can be alternatively cancelled by summing over internal $\mathcal{N}=4$ DBIVA superfields. Finally we find that $\mathcal{N}=4$ Dirac-Born-Infeld-Volkov-Akulov (DBIVA) amplitudes permit double-copy construction through two-loop order. This is shown by reproducing our unitarity based result with a loop-level double copy between $\mathcal{N}=4$ super-Yang-Mills numerators and our two-loop NLSM amplitudes. This result provides strong evidence for the existence of a color-dual representations for NLSM beyond one-loop. We conclude with an overview of how $D$-dimensional four-photon counterterms can be constructed in generality with the symmetric-structure double-copy, and outline a convenient way of counting evanescent operators using Hilbert series as generating functions. }

@@

\section{Introduction}\label{sec:intro}
Recent decades have witnessed a renaissance in our ability to compute scattering amplitudes to high orders in perturbation theory. At the epicenter of this explosion in $S$-matrix data is the unitarity method \cite{UnitarityMethod, Fusing, BDKUniarityReview}, which bypasses the standard Feynman diagram approach by constructing amplitudes directly from on-shell information. At one-loop, the unitarity method allows one to extract amplitudes by directly computing a series of unitarity cuts \cite{Forde:2007mi}. On-shell methods have furthermore unveiled novel amplitude-level structure, like the color-kinematics duality \cite{BCJ} and the double-copy construction \cite{BCJLoop}, which are completely obscured by a Lagrangian-level description of the physics. While much of the research in perturbative calculations thus far has focused on gauge theory and gravity, recent literature has investigated to what extent the $S$-matrix of effective field theory is constrained by on-shell data \cite{Adams:2006sv,Cheung:2015ota,Cheung:2016drk,Carrasco:2019yyn,Arkani-Hamed:2020blm,Carrasco:2021ptp,Chi:2021mio,Bonnefoy:2021qgu,Carrasco:2022lbm,Carrasco:2022sck,Green:2022slj,Pavao:2022kog,Chen:2022shl,Chen:2023dcx,Brown:2023srz}. In this work, we take a quantum leap to the multi-loop sector of effective field theory. As we embark on this journey, we anticipate that further probes of perturbative effective field theory data will unravel a rich tapestry of discoveries, adding to the current knowledge base and paving the way for future advancements in the field.

In this paper we study even-point (EP) effective field theory (EFT) at the multi-loop level using a combination of generalized unitarity \cite{UnitarityMethod, Fusing, BDKUniarityReview} and the double-copy construction \cite{BCJ, BCJLoop}. The EFTs that we consider are the nonlinear sigma model (NLSM) and a related family of gauge theories involving Born-Infeld theory and supersymmetric generalizations known as Dirac-Born-Infeld-Volkov-Akulov (DBIVA) theories. These families are known at tree-level to be double-copies between NLSM and Yang-Mills theories with varying amounts of supersymmetry. While we focus on these specific examples, the methods we develop can be used for perturbative calculations in any EP effective field theory.

The motivation for this work is three-fold. First, perturbative calculations in NLSM and DBIVA effective field theories are significantly simpler than the typical multi-loop calculations of renormalizable theories, like quantum-chromodynamics (QCD) and super-Yang-Mills (sYM). As we will show, the multi-loop amplitudes for these theories allow us to recycle many of the $D$-dimensional integration tools that are so powerful at one-loop order. To approach this problem, we expand on the now standard approach of Forde \cite{Forde:2007mi}, that computes one-loop amplitudes directly from 4D unitarity cuts. In \sect{sec:EMU} we'll show that there are only two basis integrals needed at two-loop four-point for EP effective field theory amplitudes. In the spirit of one-loop unitarity, the integral coefficients can be directly extracted from EMU and tensor reduction. In contrast to Forde's approach, our method of Even-point Multi-loop Unitarity (EMU) is a $D$-dimensional approach to integrand construction. This allows us to compute a large catalog of $D$-dimensional two-loop amplitudes, which provides fertile ground for cultivating insights about the multi-loop structure of effective field theories. The perturbative depth of our calculations sheds light on the exponential structure of IR divergences for NLSM amplitudes through two-loops, the anomalous behavior $U(1)$ duality invariance through two-loop order in DBIVA theories, as well as the emergence of evanescent operators relevant for anomaly cancellation in pure Born-Infeld theory. We believe these surprisingly rich physical structures could serve as a theoretical laboratory for future studies of the interplay between effective field theory operators and perturbative calculations in quantum field theory.

The second motivation is that very little is known about color-kinematics duality at multi-loop level for generic models in the web of theories \cite{BCJreview}. For a comprehensive review of color-dual representations, see refs. \cite{BCJreview, Bern:2022wqg, Adamo:2022dcm} and references therein. While there has been tremendous success of applying color-kinematics duality to $\mathcal{N}=4$ super-Yang-Mills, for which color-dual integrands are known through four-point four-loop \cite{Bern:2012uf, Neq44np, GravityFour}, five-point through three-loops, and to sEven-point at one-loop \cite{Bjerrum-Bohr:2013iza,Edison:2020uzf,Edison:2022jln}, there are many obstructions for generic gauge/gravity theories. Presently, the state-of-the-art for nonlinear sigma model (NLSM) color-dual numerators come from the $XY\!Z$ model of Cheung and Shen \cite{Cheung2016prv}, while $D$-dimensional integrands for pure Yang-Mills have only been identified through five-point one-loop \cite{He:2017spx}. At present, there are no known $Ds$-dimensional representations for either of these theories at two-loop that globally manifest the duality between color and kinematics, despite attempts in the literature \cite{OneTwoLoopPureYMBCJ, Mogull:2015adi, Bern:2015ooa, Geyer:2019hnn}. Furthermore, similar bottlenecks exists for less-than-maximal $\mathcal{N}<2$ sYM \cite{Johansson:2017bfl}. While there have been a number of recent developments in constructing manifestly color-dual Feynman rules \cite{Chen:2019ywi,Chen:2021chy,Brandhuber:2021bsf,Cheung:2021zvb,Ben-Shahar:2021zww,Cheung:2022mix, Ben-Shahar:2022ixa}, a precise definition the kinematic algebra off-shell remains elusive. Considering the known tree-level double-copy relationships between NLSM and DBIVA and higher-derivative corrections \cite{Cachazo:2014xea,Carrasco:2016ldy}, all of the amplitudes we compute should in principle participate in a double-copy formulation of our results. At the very least, the amplitudes we compute will serve as important checks on future breakthroughs in multi-loop studies of color-dual representations. In addition to traditional double-copy construction, we find that the quantum effective actions generated by loop effects permit a rather compact construction in terms of \textit{symmetric-structure double-copy}, which was introduced in recent work by the authors~\cite{Carrasco:2022jxn}.

@@

The outline of the paper is as follows: in \sect{sec:Review}, we will provide a review of the on-shell methods and integration techniques needed to probe the multi-loop physics studied in this paper. Then in \sect{sec:EMU}, we introduce the method of Even-point Multi-loop Unitarity (EMU) and compute the two-loop tensor integrals needed in this work. In \sect{sec:Loops}, we present our results, beginning with a warm-up calculation in \sect{sec:NLSMU} where we construct the integrands for NLSM through NNLO in the effective coupling, and compute the fully integrated amplitudes. We show that in $D=2-2\epsilon$, the leading IR divergences of the theory exponentiate. With the scaffolding of $D$-dimensional integration in hand, we then compute two-loop amplitudes for $\mathcal{N}=4$ DBIVA theory and pure-photon Born-Infeld theory in \sect{sec:DBIU}. There we compute the anomalies present beyond one-loop order.  Then, using the NLSM integrands of \sect{sec:NLSMU}, we perform a multi-loop double-copy to $\mathcal{N}=4$ DBIVA observables with the color-dual basis numerators available in the literature in \sect{sec:DBIvDC}. After computing the catalog of two-loop amplitudes, we study the construction of quantum effective actions in \sect{sec:Actions} as they relate to the anomalous matrix elements present in two-loop Born-Infeld amplitudes. We demonstrate in \sect{sec:Anomalies} that the one-minus anomaly at two-loop requires the introduction of an evanescent operator at $\mathcal{O}(\alpha'^4)$. Then in \sect{sec:ActionDC} we take the opportunity to discuss the application of symmetric-structure double copy to cancel these anomalies, along with their relationship to higher-spin modes as studied in a recent work by the authors. Finally, in \sect{sec:EOpsCounting} we lay out a Hilbert series framework for counting evanescent operators at general orders in mass-dimension. To conclude, in \sect{sec:Discussion} we discuss many directions of future work and summarize the insights gained from this study.

\section{Review} 
\label{sec:Review}
\subsection{Color-dressed and ordered amplitudes}
Here we provide an overview of the amplitudes nomenclature and organizational principles we use throughout the work. When working with scattering amplitudes in an on-shell framework it is convenient to introduce the following decomposition at general multiplicity, $n$, and loop order, $L$,
\be\label{fullAmp}
\mathcal{A}_{n,L} = \int \prod_{i=1}^L\frac{d^D l_i}{(2\pi)^D} \sum_{g} \frac{1}{S_g}\frac{\mathcal{N}_g}{D_g}
\ee
Throughout the work we will call, $\mathcal{A}$, the \textit{color-dressed} amplitudes, as they are the full amplitudes one would obtain from Feynman diagrams of a given gauge theory, like NLSM. As written above, $S_g$ are the internal symmetry factors for a Feynman diagram, $g$, the propagator structure is captured by the denominator function, $D_g$, and the theory-dependent interaction vertices determine numerator functions, $\mathcal{N}_g$. As written the numerator functions carry all the color and kinematic information derived from the Feynman rules of the theory. This can include both color data like structure constants, $f^{abc}$ up, and kinematic information like momenta and polarizations, depending on the particle content of the theory.

For gauge theories, it is often useful to further decompose the full color-dressed amplitudes into kinematic building blocks where the color information is stripped away. For example, at tree level, gauge theories $n$-point amplitudes can be re-expressed in terms of a trace basis decomposition as follows:
\be
\mathcal{A}_{n,\text{tree}} = \sum_{\sigma \in S^{n-1}} \text{Tr}(T^1 T^{\sigma(2)}\cdots T^{\sigma(n)}) A(1,\sigma(2),...,\sigma(n))
\ee
where the sum is taken over the inequivalent orderings of group theory generator traces. A similar decomposition applies at one-loop gauge theory amplitudes, which we will describe in \sect{sec:Loops}. The color-ordered functions are called \textit{partial amplitudes}, as they only contain on-shell information for a particular color ordering. However, like full color-dressed amplitudes, they factor on poles to lower-point partial amplitudes, and when describe vector theories they are independently gauge invariant since they weight linearly independent color traces.

When expressed in this form, many gauge theories reveal hidden redundancy that is obscured by the color-dressed amplitude formulation of \eqn{fullAmp}. When expressed in this form, the partial amplitudes for large catalog \cite{BCJreview} of gauge theories, including Yang-Mills and NLSM, are linearly dependent. The first such set of partial amplitude relations identified by Kleiss and Kluijf \cite{KK}, known formally as KK relations, relates amplitudes different color orderings and signature:
\be\label{KK}
A(1,\alpha,n,\beta)=(-1)^{|\alpha|}\sum_{\sigma\in \alpha \shuffle \beta^T} A(1,\sigma,n)
\ee
where $\beta^T$ denotes the inverse ordering of $\beta$, and $\shuffle$ is the shuffle product that enumerates ordered permutations of $\alpha$ and $\beta^T$. More recently, Bern, Johansson and one of the authors, further identified a set of kinematic relations that reduced the size of $n$-point partial amplitude basis down to $(n-3)!$ via BCJ relations \cite{BCJ}:
\be\label{BCJ}
\sum_{i=2}^{n-1}k_1\cdot(k_2+\cdots +k_i) A(2,...,i,1,...,n)=0
\ee
Underlying both sets of amplitude relations are hidden graphical organization principles that manifest the amplitude level redundancy of \eqn{KK} and \eqn{BCJ}, which we will describe in the next section. Before moving on, we emphasize that everything described thus far applies in arbitrary spacetime dimensions. As such, the aforementioned nomenclature and organization is well suited for constructing loop integrands needed for dimensional regularization.
\label{subsec:AmpReview}
